---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

1. Code for reading in the dataset and/or processing the data

```{r preprocess, results="hide"}
library(readr)
library(ggplot2)
library(dplyr)

activityFrame <- read_csv("./activity.zip")
# convert the interval column to a set of factors
activityFrame$interval <- factor(activityFrame$interval)
```


## What is mean total number of steps taken per day?

### Total steps per day:

```{r dailySteps}
dailySteps <- activityFrame %>% group_by(date) %>% summarize(dailyTotal = sum(steps, na.rm = TRUE))
print(dailySteps)
```

### Histogram showing total steps per day

2. Histogram of the total number of steps taken each day

```{r histo_dailySteps}
ggplot(dailySteps, aes(x = as.Date(date, format = "%y-%m-%d"), y = dailyTotal)) +
  geom_bar(stat = "identity") +
  labs(x = "Date", y = "Number of Steps")
```

### Mean number of steps per day

3a. Mean number of steps taken each day

```{r mean_numberSteps}
averageStepsPerDay <- mean(dailySteps$dailyTotal)
print(averageStepsPerDay)
```

### Median number of steps per day

3b. Median number of steps taken each day


```{r media_numberSteps}
medianStepsPerDay <- median(dailySteps$dailyTotal)
print(medianStepsPerDay)
```


## What is the average daily activity pattern?

### Time series plot of average number of steps taken across all days for each interval

4. Time series plot of the average number of steps taken

```{r timeSeries_Days}
intervalAverageSteps <- activityFrame %>% group_by(interval) %>% summarize(averageSteps = mean(steps, na.rm = TRUE))
ggplot(intervalAverageSteps, aes(x = interval, y = averageSteps, group = 1)) + geom_line()
```

### Time of day that contains the most number of steps (on average)

5. The 5-minute interval that, on average, contains the maximum number of steps

```{r mostNumberSteps}
sortedStepsAverage <- arrange(intervalAverageSteps, desc(averageSteps))
print(as.character(sortedStepsAverage$interval[[1]]))
```

## Imputing missing values

### Total number of missing values in dataset

```{r numberMissingValues}
sum(is.na(activityFrame$steps))
```

### Fill in missing values

6. Code to describe and show a strategy for imputing missing data

Will fill in missing values with the average for that interval as calculated earlier in #4

```{r imputeValues}
# create a copy of the frame to store the imputed values
imputedActivityFrame <- data.frame(activityFrame)
numberOfRecords <- nrow(imputedActivityFrame)
for(index in 1:numberOfRecords) {
    # if the current row contains missing data for the number of steps, let's fix that
    if(is.na(imputedActivityFrame$steps[index])) {
        # get the current interval
        intervalValue <- imputedActivityFrame$interval[index]
        # get the index into the intervalAverageSteps table
        averageIntervalIndex <- which(intervalAverageSteps$interval == intervalValue)
        # get the average steps for the interval
        averageValue <- intervalAverageSteps$averageSteps[averageIntervalIndex]
        # need to change the NA to the average value of the interval
        imputedActivityFrame$steps[index] <- averageValue
    }
}

print(head(imputedActivityFrame))
```

### Histogram of total steps taken each day (imputed data)

7. Histogram of the total number of steps taken each day after missing values are imputed

```{r histo_imputedData}
dailyTotalSteps <- imputedActivityFrame %>% group_by(date) %>% summarize(totalSteps = sum(steps, na.rm = TRUE))
ggplot(dailyTotalSteps, aes(x = as.Date(date, format = "%y-%m-%d"), y = totalSteps)) +
  geom_bar(stat = "identity") +
  labs(x = "Date", y = "Number of Steps")
```

### Mean number of steps per day (imputed data)

```{r mean_numberSteps_imputed}
averageStepsPerDay <- mean(dailyTotalSteps$totalSteps)
print(averageStepsPerDay)
```

### Median number of steps per day (imputed data)

```{r median_numberSteps_imputed}
medianStepsPerDay <- median(dailyTotalSteps$totalSteps)
print(medianStepsPerDay)
```

## Are there differences in activity patterns between weekdays and weekends?

### Create new factor variable for "weekend" or "weekday"

Let's use the dataset that contains imputed values

```{r dayType}
# create column with the weekdays to begin with
dayTypeSteps <- mutate(imputedActivityFrame, DayType = weekdays(date))
# set the day type as 'Weekend' or 'Weekday' depending on the day
dayTypeSteps$DayType[dayTypeSteps$DayType == "Saturday"] <- "Weekend"
dayTypeSteps$DayType[dayTypeSteps$DayType == "Sunday"] <- "Weekend"
dayTypeSteps$DayType[dayTypeSteps$DayType == "Monday"] <- "Weekday"
dayTypeSteps$DayType[dayTypeSteps$DayType == "Tuesday"] <- "Weekday"
dayTypeSteps$DayType[dayTypeSteps$DayType == "Wednesday"] <- "Weekday"
dayTypeSteps$DayType[dayTypeSteps$DayType == "Thursday"] <- "Weekday"
dayTypeSteps$DayType[dayTypeSteps$DayType == "Friday"] <- "Weekday"
# convert 'Weekend' and 'Weekday' to factors
dayTypeSteps$DayType <- as.factor(dayTypeSteps$DayType)
str(dayTypeSteps)
head(dayTypeSteps)
```

### Daily average for Weekend and Weekday Plot

8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r panelPlot}
intervalDayTypeSteps <- dayTypeSteps %>% group_by(interval, DayType) %>% summarize(averageSteps = mean(steps, na.rm = TRUE))
ggplot(intervalDayTypeSteps, aes(x = interval, y = averageSteps, group = DayType)) + geom_line() + facet_grid(DayType ~ .)
```
